Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0

### 2019-05-31 manuscript.Rmd

## The GTEx gene expression resource lacks eye-specific tissues
The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a comprehensive and user-friendly web app (https://gtexportal.org/); however eye tissues have not been included [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017]. Recently Ratnapriya et al. have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link. This data, as of May 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019]. The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018]. To date, no curation of the sample level metadata has been done, therefore it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed. More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016]. However iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only. We thus aimed our efforts at developing an easily accessible and reliable RNA-seq based transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.

## The eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017 at version 0.6, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018]. This interactive web app allows for quick transcript and gene comparisons across many eye tissues and dozens of other body tissues. The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to manually assess quality control for the samples. The interactive nature of some of the steps precluded efficient and regular data updates for the data.

@@

## Quality control 
We first removed samples with a Salmon calculated mapping rate less than 40%. This value was selected as being the far left tail of the distribution of mapping rates across samples (`r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`). We further filtered samples by calculating the median gene expression for each sample and remove outlier samples (data not shown) that had a median count <2 across all genes. We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count >200 across all samples. To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR R package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].

@@

dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
tsne_50$Cluster <- dbscan_cluster$cluster
col1 = 'X1'
col2 = 'X2'
# create label points for each cluster
cluster_centers <- tsne_50 %>%
  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))

>>>>

## The GTEx gene expression web app lacks eye-specific tissues
The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a comprehensive and user-friendly web app (https://gtexportal.org/); however eye tissues have not been included [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017]. Recently Ratnapriya et al. have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link. This data, as of May 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019]. The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018]. To date, no curation of the sample level metadata has been done, therefore it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed. More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016]. However iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only. We thus aimed our efforts at developing an easily accessible and reliable RNA-seq based transcriptome database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues.

## The eyeIntegration app interactively serves huge GTEx and human eye tissue datasets (EiaD)
The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017 at version 0.6, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018]. This interactive web app allows for quick transcript and gene comparisons across many eye tissues and dozens of other body tissues. The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to manually assess quality control for the samples. The interactive nature of some of the steps precluded efficient and regular data updates for the data.

@@

## Quality control 
We first removed samples with a Salmon calculated mapping rate less than 40%. This value was selected as being the far left tail of the distribution of mapping rates across samples (`r supFig_cap(name = 'mapping_rate_cap', display = 'cite')`). We removed lowly expressed genes by calculating the median expression across all samples for each gene and kept genes that had a median count >200 across all samples. To reduce the noise from experimental variability between each study, we normalized samples by sequence library size using the calcNormFactors function from the edgeR R package, and then quantile smoothed expression data using the R package qsmooth at the tissue level [@hicksSmoothQuantileNormalization2018a; @robinsonEdgeRBioconductorPackage2010].

@@

tsne_50 <- tsne_50 %>% filter(Tissue != 'Choroid Plexus', !(Tissue == 'Retina' & X2 > 20))
dbscan_cluster <- dbscan(tsne_50 %>% select(X1, X2),  minPts = 3, eps = 2.5)
tsne_50$Cluster <- dbscan_cluster$cluster
col1 = 'X1'
col2 = 'X2'
# create label points for each cluster
cluster_centers <- tsne_50 %>%
  left_join(.,core_tight_2019) %>% group_by(Cluster) %>%
  summarise(C1=mean(!!sym(col1)),C2=mean(!!sym(col2)),Tissue=paste(unique(Tissue),collapse=','))
